---
title: "Get Started"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

ine_is_reachable <- tryCatch({
  con <- suppressWarnings(socketConnection("www.ine.pt", port = 443, timeout = 2))
  close(con)
  TRUE
}, error = function(e) FALSE)
```

# Installation

You can install the development version of `MobilityDataPT` from GitHub with:
```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("U-Shift/MobilityDataPT")
```

# Load the package

```{r eval=FALSE}
library(MobilityDataPT)
```

# Key functions

## 1. Administrative Boundaries & Postal Codes

Access Eurostat GISCO postal code datasets and spatial coordinates for Portugal or other EU countries:

```{r}
# Download and access full postal code database for Portugal
db <- MobilityDataPT::postal_code_database(cntr_id = "PT", year = 2024, crs = 4326)
db |>
  dplyr::select(POSTCODE, LAU_NAME, Shape) |>
  dplyr::sample_n(5)

# Get spatial coordinates for specific postal codes
coords <- MobilityDataPT::get_postal_code_coordinates(c("1000-001", "2800-001"))
coords
```

## 2. Census Origin-Destination (OD) Data

Retrieve INE Census 2021 Origin-Destination data for home-to-work and home-to-study mobility flows:

```{r eval = ine_is_reachable}
# Fetch available region filtering options
regions <- MobilityDataPT::census_od_regions()
head(regions)

# Fetch OD mobility records for a specific region ID (e.g., "PT" or "0603")
od_data <- MobilityDataPT::census_od(id = "PT")
head(od_data)
```

## 3. Road & Itinerary Toll Costs

Calculate expected toll costs across vehicle classes (C1, C2, C3, C4) for spatial route itineraries using the Infraestruturas de Portugal API service:

```{r}
sf_itinerary <- sf::st_read(system.file("extdata/samples",
  "tool_itinerary_25AbrilBridge.gpkg",
  package = "MobilityDataPT"
), quiet = TRUE)

# Calculate toll costs
tolls <- MobilityDataPT::tools_for_itinerary(sf_itinerary)
tolls
```